Tidal streams in triaxial systems

### 2015-03-31 ms.tex
figures at end, adding to potential section

% commands
\newcommand{\given}{\,|\,}
\newcommand{\dd}{\mathrm{d}}
\newcommand{\transpose}[1]{{#1}^{\mathsf{T}}}
\newcommand{\inverse}[1]{{#1}^{-1}}
\newcommand{\msun}{\mathrm{M}_\odot}
\newcommand{\bs}[1]{\boldsymbol{#1}}
\newcommand{\ident}{\mathbb{1}}
\newcommand{\inttime}{t_{\rm int}}
\newcommand{\fdrate}{\mathcal{R}_\Omega}

@@

%\begin{figure*}[!h]
%\begin{center}
%\includegraphics[width=\textwidth]{PATH}
%\caption{ CAPTION }\label{fig:FIGNAME}
%\end{center}
%\end{figure*}

@@

As a satellite galaxy or globular cluster orbits within some larger system, mass is eroded due to the tidal forces of the host galaxy potential. If the orbit of such an object is regular, only mildly eccentric, and stays far from sharp density features (e.g., the disk), the disruption is a steady process and the mass is slowly stretched into long streams of matter: tidal (stellar) streams. The phase-space density and therefore the observable properties of these tidal streams are sensitive to the internal progenitor properties, orbit of the progenitor, and  mass distribution of the host galaxy. The morphological evolution of the debris depends on the spread of orbital properties (e.g., actions or frequencies) of the debris and the orbit of the progenitor system, both of which are also set by the shape and radial profile of the gravitational potential of the host galaxy. In the halo of the Milky Way, tidal streams of stars are found over a large range of distances ($\sim$10-100~kpc). By simultaneously modeling the observed phase-space density of stream stars along with the potential, it is hoped that we may infer the dark matter distribution of the Milky Way on these scales.

@@

The obvious difference between the two potentials considered by \citet{pearson15} is the extra symmetry of the oblate potential. It is well known that the number of symmetries or the number of degrees of freedom of a potential plays a major role in determining the orbital structure within the potential. The peculiar shape of the particular triaxial potential also likely has significant chaotic regions within the potential. \citet{pearson15} tested the stochasticity of the orbit of the fanned debris progenitor by computing the Lyapunov exponent but found that the orbit is consistent with being regular over dynamically relevant timescales (many Hubble times). It has been shown qualitatively that along some strongly chaotic orbits, tidal streams form large, diffuse fans of debris \citep[e.g.,][]{fardal14}. However, it is unknown how the resultant properties of the debris (e.g., density or length of the stream) depend on the degree of stochasticity; the result from \citet{pearson15} suggests that even a weakly chaotic progenitor orbit effects the density and therefore observability of tidal streams.

@@

The density distributions within dark matter halos formed in cosmological N-body simulations are typically triaxial \citep[e.g.,][]{jing02, bett07, zemp09, veraciro11}. \citet{jing02} found that a triaxial generalization of the classic NFW density profile \citep{navarro96} generates excellent fits to their high-resolution N-body simulations, and they provide probability distributions for the axis ratios of the many halos formed in their simulations. They find median axis ratios of $c/a \approx 0.55$ and $b/a \approx 0.77$ where $a$ is the major axis, $b$ the intermediate, and $c$ the minor axis.\footnote{Note that \citet{jing02} use the opposite notation so that $c$ is the major and $a$ is the minor axis.} These are consistent with constraints from weak lensing that place a lower limit on minor-to-major axis ratios of $c/a\gtrsim0.5$ \citep{vanuitert12}. They find significant scatter in the distributions of concentration parameter, $c_e$, or scale radius (depending on choice of parametrization).

@@

\begin{figure*}[!h]
\begin{center}
\includegraphics[width=0.75\textwidth]{figures/potential.pdf}
\caption{Rotation curve along $x$, $y$, and $z$ axes for the triaxial, NFW potential } \label{fig:potential}
\end{center}
\end{figure*}

This potential is a simple and unrealistic model for the total potential of a Milky-Way-like galaxy, however it represents a conservative choice for exploring the effect of chaos in the halos of such galaxies. Realistic Galactic potentials will have a significant component due to the disk and bulge and may have twisting inertia tensors (cite), radially changing axis ratios \citep[e.g.,][]{kazantzidis04,debattista08,veraciro11}, significant substructure \citep{moore98,zemp09}, or time dependence \citep[either from bulk rotation, mass growth, mergers, etc.][]{bailin05}. We expect all of these effects to increase the amount and influence of chaos; see the Discussion (Section~\ref{sec:discussion}) for a few simple demonstrations.

@@

\begin{figure*}[!h]
\begin{center}
\includegraphics[width=0.6\textwidth]{figures/orbit-tree.pdf}
\caption{A rough classification of the types of orbits found in triaxial potentials. Regular orbits are classified by the number of resonance relations obeyed: conditionally periodic (\emph{none}), uni-resonant (\emph{one}), bi-resonant (\emph{two}). Irregular or chaotic orbits are simply separated into two classes: strongly and weakly chaotic, defined by the timescales of interest in a problem. When the characteristic chaotic timescale is much longer than, for example, the age of the universe, the orbit is only weakly chaotic. } \label{fig:orbit-tree}
\end{center}
\end{figure*}

@@

\begin{figure*}[!h]
\begin{center}
\includegraphics[width=0.5\textwidth]{figures/log-freqmap.png}
\caption{Ratio of cartesian frequencies for a grid of iso-energy box orbits integrated in the potential of Equation~\ref{eq:logpotential}. Stable resonances appear as linear over-densities and unstable resonances appear as linear under-densities. The regularity of the points in this map reflects the input grid of initial conditions. Points that appear to be erratically scattered are chaotic orbits where the frequencies are changing with time.} \label{fig:logfreqs}
\end{center}
\end{figure*}

@@

\begin{figure*}[!h]
\begin{center}
\includegraphics[width=0.9\textwidth]{figures/two-orbits.pdf}
\caption{ Two orbits integrated in the triaxial potential described in Section~\ref{sec:potential}. Left panels (black) show two consecutive sections of orbit, where the integration time for each section (rows) is set to 100 times the maximum orbital period. Right panels (red) shows the same for the chaotic orbit. The appearance of the orbit in configuration space changes appreciably over timescales of tens of orbital periods. } \label{fig:two-orbits}
\end{center}
\end{figure*}

@@

\begin{figure*}[!h]
\begin{center}
\includegraphics[width=0.62\textwidth]{figures/lyap-orbits.pdf}
\caption{ FTMLE (Equation~\ref{eq:ftmle}) computed over 1000 orbital periods (the maximum orbital period) for Orbit A (left) and Orbit B (right). For Orbit A, the FTMLE asymptotes towards zero with a characteristic power-law slope. For Orbit B, the FTMLE begins to saturate to a constant; the Lyapunov exponent estimated for this orbit is $\lambda_{\rm max} \approx 0.07 \, {\rm Gyr}^{-1}$ corresponding to a Lyapunov time of $t_\lambda \approx 14 \, {Gyr}$. Over-plotted on both panels are lines (dashed, blue) that show $t^{-0.9}$ behavior.} \label{fig:lyap-orbits}
\end{center}
\end{figure*}
\begin{figure*}[!h]
\begin{center}
\includegraphics[width=0.62\textwidth]{figures/freq-diff-orbits.pdf}
\caption{ Difference in frequencies computed for successive sections of 100 orbital periods from the frequencies computed in the initial section for Orbit A (left) and Orbit B (right). The three lines in each panel are for the three fundamental frequencies. Dashed (red) horizontal line shows the estimate of the uncertainty for the frequencies computed for Orbit A. After many hundreds of orbital periods, the frequency deviation is not appreciable indicating again that the orbit is regular or only very weakly chaotic. The frequency deviations computed for Orbit B, however, show a clear increasing trend, indicating that the frequencies are diffusing with time. } \label{fig:fdiff-orbits} % The frequency diffusion rate defined in Equation~\ref{eq:fdrate} measures the initial slope of this trend.
\end{center}
\end{figure*}

@@

\begin{figure*}[!h]
\begin{center}
\includegraphics[width=0.75\textwidth]{figures/D_KL.pdf}
\caption{This is really just a placeholder that shows what I want to show here...} \label{fig:DKL}
\end{center}
\end{figure*}

@@

\begin{figure}[!h]
\centering
	\includegraphics[width=0.4\columnwidth] {figures/lyap_map_xz.png}
	\caption{Not the final figures. Need to re-make with same axis limits, remove labels, put in same PDF file...etc.} 
	\label{fig:lyapmap}
\end{figure}

@@

\begin{figure}[!h]
\centering
	\includegraphics[width=0.4\columnwidth] {figures/kld-map.png}
	\caption{Not the final figures. Need to re-make with same axis limits, remove labels, put in same PDF file...etc.} 
	\label{fig:kldmap}
\end{figure}

@@

\bibliographystyle{apj}
\bibliography{refs}

>>>>

% commands
\newcommand{\given}{\,|\,}
\newcommand{\dd}{\mathrm{d}}
\newcommand{\transpose}[1]{{#1}^{\mathsf{T}}}
\newcommand{\inverse}[1]{{#1}^{-1}}
\newcommand{\msun}{\ensuremath{\mathrm{M}_\odot}}
\newcommand{\bs}[1]{\boldsymbol{#1}}
\newcommand{\ident}{\mathbb{1}}
\newcommand{\inttime}{t_{\rm int}}
\newcommand{\fdrate}{\mathcal{R}_\Omega}

@@

@@

As a satellite galaxy or globular cluster orbits within some larger system, mass is eroded due to the tidal forces of the host galaxy potential. If the orbit of such an object is regular, only mildly eccentric, and stays far from sharp density features (e.g., the disk), the disruption is a steady process and the mass is slowly stretched into long streams of matter: stellar tidal streams. The phase-space density and therefore the observable properties of these tidal streams are sensitive to the internal progenitor properties, orbit of the progenitor, and  mass distribution of the host galaxy. The morphological evolution of the debris depends on the spread of orbital properties (e.g., actions or frequencies) of the debris and the orbit of the progenitor system, both of which are also set by the shape and radial profile of the gravitational potential of the host galaxy. In the halo of the Milky Way, tidal streams of stars are found over a large range of distances ($\approx$10-100~kpc). By simultaneously modeling the observed phase-space density of stream stars along with the potential, it is hoped that we may infer the dark matter distribution of the Milky Way on these scales.

@@

The obvious difference between the two potentials considered by \citet{pearson15} is the extra symmetry of the oblate potential. It is well known that the number of symmetries or the number of degrees of freedom of a potential plays a major role in determining the orbital structure within the potential. The peculiar shape of the particular triaxial potential also likely has significant chaotic regions within the potential. \citet{pearson15} tested the stochasticity of the orbit of the fanned debris progenitor by computing the Lyapunov exponent but found that the orbit is consistent with being regular over dynamically relevant timescales (many Hubble times). It has been shown qualitatively that along some strongly chaotic orbits, tidal streams form large, diffuse fans of debris \citep[e.g.,][]{fardal14}. However, it is unknown how the resultant properties of the debris (e.g., density or length of the stream) depend on the degree of stochasticity; the result from \citet{pearson15} suggests that even a weakly chaotic progenitor orbit affects the density and therefore observability of tidal streams.

@@

The density distributions within dark matter halos formed in cosmological N-body simulations are generically triaxial \citep[e.g.,][]{jing02, bett07, zemp09, veraciro11}. With the inclusion of baryonic physics and sub-grid prescriptions for energy input due to supernovae and other feedback mechanisms, the inner potential ($\lesssim$$0.1R_{\rm vir}$ for a $\approx$$10^{12}~\msun$ halo mass) typically becomes more spherical, though the magnitude of this reshaping depends on the particular merger history and star formation efficiency within a given halo \citep[e.g.,][]{dubinski??, recent NiHao stuff}. It is less clear what happens to the outer halo where tidal streams are more readily found; for this reason, we ask what we would expect to happen to tidal streams if the halo in which they form is strongly triaxial.

 \citet[][hereafter JS02]{jing02} found that a triaxial generalization of the classic NFW density profile \citep{navarro96} generates excellent fits to the density distributions within haloes in their high-resolution N-body simulations, and they provide probability distributions for the axis ratios of a large sample of these halos. JS02 finds median axis ratios of $c/a \approx 0.55$ and $b/a \approx 0.77$ where $a$ is the major axis, $b$ the intermediate, and $c$ the minor axis.\footnote{Note that JS02 use the opposite notation so that $c$ is the major and $a$ is the minor axis.} These are largely consistent with findings from more recent simulations \citep[e.g.,][]{??, NiHao} and consistent with constraints from weak lensing that place a lower limit on minor-to-major axis ratios of $c/a\gtrsim0.5$ \citep{vanuitert12}. JS02 finds significant scatter in the distributions of concentration parameter, $c_e$, or scale radius (depending on choice of parametrization).

@@

% Figure 1
%\begin{figure*}[!h]
%\begin{center}
%\includegraphics[width=0.75\textwidth]{figures/potential.pdf}
%\caption{Rotation curve along $x$, $y$, and $z$ axes for the triaxial, NFW potential } \label{fig:potential}
%\end{center}
%\end{figure*}

This potential is a simple and unrealistic model for the total potential of a Milky-Way-like galaxy, however it represents a conservative choice for exploring the structure of orbits in the halos of such galaxies. Realistic Galactic potentials will have a significant component due to the disk and bulge and may have twisting inertia tensors (cite), radially changing axis ratios \citep[e.g.,][]{kazantzidis04,debattista08,veraciro11}, significant substructure \citep{moore98,zemp09}, or time dependence \citep[either from bulk rotation, mass growth, mergers, etc.][]{bailin05}. We expect all of these effects to increase the amount and influence of chaos; see the Discussion (Section~\ref{sec:discussion}) for a few simple demonstrations.

@@

% Figure 2
%\begin{figure*}[!h]
%\begin{center}
%\includegraphics[width=0.6\textwidth]{figures/orbit-tree.pdf}
%\caption{A rough classification of the types of orbits found in triaxial potentials. Regular orbits are classified by the number of resonance relations obeyed: conditionally periodic (\emph{none}), uni-resonant (\emph{one}), bi-resonant (\emph{two}). Irregular or chaotic orbits are simply separated into two classes: strongly and weakly chaotic, defined by the timescales of interest in a problem. When the characteristic chaotic timescale is much longer than, for example, the age of the universe, the orbit is only weakly chaotic. } \label{fig:orbit-tree}
%\end{center}
%\end{figure*}

@@

% Figure 3
%\begin{figure*}[!h]
%\begin{center}
%\includegraphics[width=0.5\textwidth]{figures/log-freqmap.png}
%\caption{Ratio of cartesian frequencies for a grid of iso-energy box orbits integrated in the potential of Equation~\ref{eq:logpotential}. Stable resonances appear as linear over-densities and unstable resonances appear as linear under-densities. The regularity of the points in this map reflects the input grid of initial conditions. Points that appear to be erratically scattered are chaotic orbits where the frequencies are changing with time.} \label{fig:logfreqs}
%\end{center}
%\end{figure*}

@@

% Figure 4
%\begin{figure*}[!h]
%\begin{center}
%\includegraphics[width=0.9\textwidth]{figures/two-orbits.pdf}
%\caption{ Two orbits integrated in the triaxial potential described in Section~\ref{sec:potential}. Left panels (black) show two consecutive sections of orbit, where the integration time for each section (rows) is set to 100 times the maximum orbital period. Right panels (red) shows the same for the chaotic orbit. The appearance of the orbit in configuration space changes appreciably over timescales of tens of orbital periods. } \label{fig:two-orbits}
%\end{center}
%\end{figure*}

@@

% Figure 5, 6
%\begin{figure*}[!h]
%\begin{center}
%\includegraphics[width=0.62\textwidth]{figures/lyap-orbits.pdf}
%\caption{ FTMLE (Equation~\ref{eq:ftmle}) computed over 1000 orbital periods (the maximum orbital period) for Orbit A (left) and Orbit B (right). For Orbit A, the FTMLE asymptotes towards zero with a characteristic power-law slope. For Orbit B, the FTMLE begins to saturate to a constant; the Lyapunov exponent estimated for this orbit is $\lambda_{\rm max} \approx 0.07 \, {\rm Gyr}^{-1}$ corresponding to a Lyapunov time of $t_\lambda \approx 14 \, {Gyr}$. Over-plotted on both panels are lines (dashed, blue) that show $t^{-0.9}$ behavior.} \label{fig:lyap-orbits}
%\end{center}
%\end{figure*}
%
%\begin{figure*}[!h]
%\begin{center}
%\includegraphics[width=0.62\textwidth]{figures/freq-diff-orbits.pdf}
%\caption{ Difference in frequencies computed for successive sections of 100 orbital periods from the frequencies computed in the initial section for Orbit A (left) and Orbit B (right). The three lines in each panel are for the three fundamental frequencies. Dashed (red) horizontal line shows the estimate of the uncertainty for the frequencies computed for Orbit A. After many hundreds of orbital periods, the frequency deviation is not appreciable indicating again that the orbit is regular or only very weakly chaotic. The frequency deviations computed for Orbit B, however, show a clear increasing trend, indicating that the frequencies are diffusing with time. } \label{fig:fdiff-orbits} % The frequency diffusion rate defined in Equation~\ref{eq:fdrate} measures the initial slope of this trend.
%\end{center}
%\end{figure*}

@@

@@

%\begin{figure}[!h]
%\centering
%	\includegraphics[width=0.4\columnwidth] {figures/lyap_map_xz.png}
%	\caption{Not the final figures. Need to re-make with same axis limits, remove labels, put in same PDF file...etc.} 
%	\label{fig:lyapmap}
%\end{figure}

@@

%\begin{figure}[!h]
%\centering
%	\includegraphics[width=0.4\columnwidth] {figures/kld-map.png}
%	\caption{Not the final figures. Need to re-make with same axis limits, remove labels, put in same PDF file...etc.} 
%	\label{fig:kldmap}
%\end{figure}

@@

\bibliographystyle{apj}
\bibliography{refs}
\clearpage

% Figure 1
\begin{figure*}[!p]
\begin{center}
\includegraphics[width=\textwidth]{figures/potential.pdf}
\caption{Rotation curve along $x$, $y$, and $z$ axes for the triaxial, NFW potential } \label{fig:potential}
\end{center}
\end{figure*}

% Figure 2
\begin{figure*}[!p]
\begin{center}
\includegraphics[width=\textwidth]{figures/orbit-tree.pdf}
\caption{A rough classification of the types of orbits found in triaxial potentials. Regular orbits are classified by the number of resonance relations obeyed: conditionally periodic (\emph{none}), uni-resonant (\emph{one}), bi-resonant (\emph{two}). Irregular or chaotic orbits are simply separated into two classes: strongly and weakly chaotic, defined by the timescales of interest in a problem. When the characteristic chaotic timescale is much longer than, for example, the age of the universe, the orbit is only weakly chaotic. } \label{fig:orbit-tree}
\end{center}
\end{figure*}

% Figure 3
\begin{figure*}[p]
\begin{center}
\includegraphics[width=\textwidth]{figures/log-freqmap.png}
\caption{Ratio of cartesian frequencies for a grid of iso-energy box orbits integrated in the potential of Equation~\ref{eq:logpotential}. Stable resonances appear as linear over-densities and unstable resonances appear as linear under-densities. The regularity of the points in this map reflects the input grid of initial conditions. Points that appear to be erratically scattered are chaotic orbits where the frequencies are changing with time.} \label{fig:logfreqs}
\end{center}
\end{figure*}

% Figure 4
\begin{figure*}[!p]
\begin{center}
\includegraphics[width=\textwidth]{figures/two-orbits.pdf}
\caption{ Two orbits integrated in the triaxial potential described in Section~\ref{sec:potential}. Left panels (black) show two consecutive sections of orbit, where the integration time for each section (rows) is set to 100 times the maximum orbital period. Right panels (red) shows the same for the chaotic orbit. The appearance of the orbit in configuration space changes appreciably over timescales of tens of orbital periods. } \label{fig:two-orbits}
\end{center}
\end{figure*}

% Figure 5
\begin{figure*}[p]
\begin{center}
\includegraphics[width=\textwidth]{figures/lyap-orbits.pdf}
\caption{ FTMLE (Equation~\ref{eq:ftmle}) computed over 1000 orbital periods (the maximum orbital period) for Orbit A (left) and Orbit B (right). For Orbit A, the FTMLE asymptotes towards zero with a characteristic power-law slope. For Orbit B, the FTMLE begins to saturate to a constant; the Lyapunov exponent estimated for this orbit is $\lambda_{\rm max} \approx 0.07 \, {\rm Gyr}^{-1}$ corresponding to a Lyapunov time of $t_\lambda \approx 14 \, {Gyr}$. Over-plotted on both panels are lines (dashed, blue) that show $t^{-0.9}$ behavior.} \label{fig:lyap-orbits}
\end{center}
\end{figure*}

% Figure 6
\begin{figure*}[p]
\begin{center}
\includegraphics[width=\textwidth]{figures/freq-diff-orbits.pdf}
\caption{ Difference in frequencies computed for successive sections of 100 orbital periods from the frequencies computed in the initial section for Orbit A (left) and Orbit B (right). The three lines in each panel are for the three fundamental frequencies. Dashed (red) horizontal line shows the estimate of the uncertainty for the frequencies computed for Orbit A. After many hundreds of orbital periods, the frequency deviation is not appreciable indicating again that the orbit is regular or only very weakly chaotic. The frequency deviations computed for Orbit B, however, show a clear increasing trend, indicating that the frequencies are diffusing with time. } \label{fig:fdiff-orbits} % The frequency diffusion rate defined in Equation~\ref{eq:fdrate} measures the initial slope of this trend.
\end{center}
\end{figure*}

% Figure 7
\begin{figure*}[p]
\begin{center}
\includegraphics[width=\textwidth]{figures/D_KL.pdf}
\caption{This is really just a placeholder that shows what I want to show here...} \label{fig:DKL}
\end{center}
\end{figure*}

% Figure 8
\begin{figure}[p]
\centering
	\includegraphics[width=\columnwidth] {figures/lyap_map_xz.png}
	\caption{Not the final figures. Need to re-make with same axis limits, remove labels, put in same PDF file...etc.} 
	\label{fig:lyapmap}
\end{figure}

% Figure 9
%\begin{figure}[!h]
%\centering
%	\includegraphics[width=0.4\columnwidth] {figures/kld-map.png}
%	\caption{Not the final figures. Need to re-make with same axis limits, remove labels, put in same PDF file...etc.} 
%	\label{fig:kldmap}
%\end{figure}
